\documentclass{article}

\usepackage{amsmath,amssymb,blkarray}

\newcommand{\ta}{\mathtt{a}}
\newcommand{\tb}{\mathtt{b}}
\newcommand{\tn}{\mathtt{n}}
\newcommand{\tm}{\mathtt{m}}
\newcommand{\tx}{\mathtt{x}}
\newcommand{\ty}{\mathtt{y}}
\newcommand{\tE}{\mathtt{E}}
\newcommand{\tP}{\mathtt{P}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\sR}{\mathbb{R}}

\title{Interpreting attention with toy models}
\author{John MacCormick}
\date{\today}


\begin{document}

\maketitle

This document attempts to explain some aspects of the attention
mechanism in transformer neural network models. We assume the reader
is familiar with conventional definitions and treatments of the
attention mechanism. Here, we focus on some extremely simplified
variants of transformer architectures, to explain some elementary
properties of the attention mechanism that may not be obvious in fully
featured architectures.

We typically use a very small vocabulary of tokens. As an initial
example, consider the vocabulary $\bV=\{\ta, \tb, \tE, \tP\}$. In this
vocabulary and in the rest of the document: $\tE$ is the
end-of-sequence token, denoted \texttt{<EOS>} by many authors; $\tP$
is the padding token, denoted \texttt{<PAD>} by many authors; letters
such as $\ta, \tb, \tm, \tn, \tx, \ty$ denote regular tokens, which
might represent words such as \texttt{apple} or \texttt{banana} in a
more realistic application.

We write $v$ for the size of the vocabulary, i.e.\ $v=|\bV|=4$ for the
above example. Tokens in a vocabulary are assigned ID numbers starting
from 1. In our initial example the ID numbers are
$\ta\to 1, \tb\to 2, \tE\to 3, \tP\to 4.$

The input to a transformer is a sequence of tokens. We use the one-hot
embedding of tokens in $\sR^v$, so the token with ID number $i$ is
mapped to a $1 \times v$ row vector whose $i$th coordinate is 1 and
remaining coordinates are zeroes. For example, token $\tE$ maps to
$(0,0,1,0)$. We use $n$ to denote the number of tokens in an input
sequence. A sequence of $n$ tokens maps (i.e. embeds) to a
$n \times v$ \emph{embedded matrix}, denoted $X$. For example, the
input $s = \mathtt{baE}$ embeds as
\begin{displaymath}
  X = X(s)=
\begin{pmatrix}
0 & 1 & 0 & 0 \\
1 & 0 & 0 & 0 \\
0 & 0 & 1 & 0 
\end{pmatrix}.
\end{displaymath}
As shown in this equation, we often omit the dependence of $X$ on $s$.

We will often label matrix rows and columns with corresponding
tokens. For example, the $X$ above can be written with column labels
in order of token ID and row labels in order of input token, as
follows:
\begin{displaymath}
  X = 
    \begin{blockarray}{ccccc}
       & \ta & \tb & \tE & \tP  \\
      \begin{block}{c(cccc)}
        \tb & 0 & 1 & 0 & 0 \\
        \ta & 1 & 0 & 0 & 0 \\
        \tE & 0 & 0 & 1 & 0 \\
      \end{block}
    \end{blockarray}
\end{displaymath}

In our toy system, the amount of attention between ordered pairs of
tokens is specified by a $v \times v$ \emph{relationship matrix}
$R$. (Most authors do not define $R$ directly, instead deriving it
from query and key matrices $W_q, W_k$. We will incorporate query and
key matrices in our framework below, but the simplest toy model is
best defined in terms of $R$ directly.)
% The relationship matrix $R$ has the following properties:
% \begin{itemize}
% \item Element $R_{ij}$ conveys the amount of attention paid by token $i$ to token $j$, with higher values implying greater attention.
% \item Entries of $R$ use units that scale logarithmically:
%   $R_{ij}=-\infty$ implies zero attention from $i$ to $j$; any finite
%   value of $R_{ij}$ (including negative or zero) implies nonzero
%   attention; adding a fixed increment to $R_{ij}$ increases the
%   implied attention by a multiplicative factor.
% \end{itemize}
% The logarithmic units are due to the fact that a softmax function will
% be applied later, during the attention computation. The softmax
% function involves exponentiation, which will transform $-\infty$ to 0
% and any finite value (including negative or zero) to a positive value.
The relationship matrix $R$ has the following properties:
\begin{itemize}
\item Element $R_{ij}$ conveys the amount of attention paid by token $i$ to token $j$, with higher values implying greater attention.
\item Entries of $R$ are nonnegative.
\item Rows of $R$ sum to 1, so the total amount of attention paid by a
  given token to all other token sums to 1.
\end{itemize}
Here is an example of a possible relationship matrix for the
vocabulary $\bV$ given above:
\begin{displaymath}
  R = 
    \begin{blockarray}{ccccc}
       & \ta & \tb & \tE & \tP  \\
      \begin{block}{c(cccc)}
        \ta & 0 & 1 & 0 & 0 \\
        \tb & 0.5 & 0.5 & 0 & 0 \\
        \tE & 0.25 & 0.25 & 0.25 & 0.25 \\
        \tP & 0 & 0 & 0 & 1 \\
      \end{block}
    \end{blockarray}
\end{displaymath}
The pairwise attention of tokens in an input string is defined via
$R$. Specifically, given input sequence $s$ of length $n$, we first
calculate the embedded matrix $X\equiv X(s)$, then define the
$n \times n$ attention matrix $A$ by
\begin{displaymath}
  A = A(s) = A(X) = X R\, X^T
\end{displaymath}


\end{document}
